Sara Benist MADA Project

### 2023-04-07 code/analysis_code/statistical_analysis.qmd
Part 4 of project
updated references, machine learning stat analysis, inserted figures into manuscript

```{r, loaddata}
```
# Model fitting
Now that the data has been explored, I would like to start looking at models for the main outcomes and predictors of the data set. First, we will focus split the data into testing and training sets and set up the cross-validation folds. The training set will be used to tune the model parameters and the testing set will be used to test model performance on new data.

>>>>

```{r, loaddata}
```

# Basic modelling
For some basic modeling, I will be fitting linear models to the TB outcomes with the subgroups as predictors. This will offer another view of the data. The prediction model also offers a way to explore the data by submitting a sample prediction of TB incidence for females and males in India.
```{r, basicmodels}
```
The biggest issue with the linear modeling is the assumption that the relationship between predictors and outcomes is linear, and the model cannot be tuned and replicated.

# Model fitting
Now that the data has been explored, I would like to start looking at models for the main outcomes and predictors of the data set. First, we will focus split the data into testing and training sets and set up the cross-validation folds. The training set will be used to tune the model parameters and the testing set will be used to test model performance on new data.

### 2023-04-07 products/manuscript/Manuscript.qmd
Part 4 of project
updated references, machine learning stat analysis, inserted figures into manuscript

@@

Tuberculosis is an infectious disease caused by Mycobacterium tuberculosis and is the leading infectious cause of death in the world even though the disease is both treatable and curable[@WHO2022]. Approximately 10 million people are infected each year, and 1.5 million die from the disease [@WHO2022]. Humanity has been impacted by TB for centuries, but the burden of disease is not equal across countries and populations [@WHO2022b]. The World Health Organization marked 30 countries as having highest burden of TB due to the number of cases, presence of multi-drug resistance strains, and the high mortality within certain populations such as HIV [@WHO2021b]. To understand more about why some countries or groups of people experience the higher burden, the World Health Organization gathered data on several health equity indicators subgroups. This project will explore and analyze the disparities within the indicators and the effect on tuberculosis outcomes.

@@

## Questions/Hypotheses to be addressed

@@

In addition to the inequality and indicator predictors provided with the dataset, I would like to examine differences in TB outcomes based on level of TB burden. I would need to add another classification based on the literature to indicate what levels of TB incidence and prevalence would fall into each level of disease burden.

@@

To analyze, I would need to determine how to handle the missing data. The years between countries are not specific, and not all indicators were collected for each country. At the moment, I would subset the data based on indicator and remove incomplete observations to analyze each indicator individually. For the final analysis, I will most likely focus on TB indicators that are significantly different between subgroups or complete enough for further analysis. As the class progresses, I look forward to learning other analysis techniques. Eventually, I would like to create a dataset that can produce the statistical information, plots, and models that explore TB outcomes based on inequality measures, TB indicators, and/or level of burden.

@@

For the statistical analysis of the data, I created linear models for the main outcomes (TB incidence, TB mortality, and TB prevalence) based on the subgroups of the indicator. An example is shown below. This method was repeated for the `wide_data` and `highburden` countries.

@@

Other indicators were also analyzed and fit to a linear model as shown below.

@@

In the next step of my project, I would like to manipulate the data set so I can use the social determinant indicators as predictors for the main outcomes. This will most likely be limited to the high burden countries.

@@

*(Part 2 of project: exploratory analysis)* Please see `exploratory_analysis.qmd` for full code describing data exploratory analysis.

@@

Most indicators showed the greatest disparities due to economic status and education level. Further exploratory analysis should look into level of TB burden as a possible variable that should be considered, and statistical analysis in the next step should attempt to quantify the disparities between subgroups and the impact on TB outcomes.

## Basic statistical analysis

*(Part 3 of project: linear model fitting and predictions, more to be added)* TBA: simple statistical models for associations in data

@@

TBA: final analysis with statistic/ML methods, meaningful figures/tables, load saved products here

>>>>

# Questions for Dr. Handel
I would like to be able to predict the main TB outcomes based on the health equity indicators (ex: level of TB incidence predicted from % of people with drug-resistant strains), but I cannot join the rows together correctly. When I tried bind_rows() on the one of the countries to see if it would work, the outcome rows are still separated and if I remove year to force everything onto the same row, I have 90 variables and they are grouped within the same cell. I added an example below. Do you have any advice?

```{r}
#path to data
data_location <- here::here("data","raw_data","202206-repository-tb.xlsx")

#without year removed: outcomes remain on a separate row
test <- rawdata %>% 
  select(c(setting, year, indicator_name,
           dimension, subgroup,
           estimate)) %>% 
  pivot_wider(names_from = c("indicator_name", "dimension", "subgroup"), values_from = "estimate") %>% 
  filter(setting == "Afghanistan") %>% bind_rows()
test

#with year removed: cells contain multiple observations
test2 <- rawdata %>% 
  select(c(setting, indicator_name,
           dimension, subgroup,
           estimate)) %>% 
  pivot_wider(names_from = c("indicator_name", "dimension", "subgroup"), values_from = "estimate") %>% 
  filter(setting == "Afghanistan") %>% bind_rows()
test2
```
If it is not feasible, I will be analyzing TB outcomes separately from indicators instead.

@@

Tuberculosis is an infectious disease caused by *Mycobacterium tuberculosis* and is the leading infectious cause of death in the world even though the disease is both treatable and curable[@WHO2022]. Approximately 10 million people are infected each year, and 1.5 million die from the disease [@WHO2022]. Humanity has been impacted by TB for centuries, but the burden of disease is not equal across countries and populations [@WHO2022b]. The World Health Organization marked 30 countries as having highest burden of TB due to the number of cases, presence of multi-drug resistance strains, and the high mortality within certain populations such as HIV [@WHO2021b]. To understand more about why some countries or groups of people experience the higher burden, the World Health Organization gathered data on several health equity indicators subgroups. The Global Tuberculosis Programme houses the End TB Strategy with a goal to eradicating TB. Their current objective is to reduce TB incidence by 80% and TB deaths by 90% by 2030 [@WHO2022c]. Previous research has focused on trends in TB outcomes,; this project will explore and analyze the disparities within the indicators based on subgroup and burden of disease as well as analyizing trends in tuberculosis outcomes.

@@

## Questions/Hypotheses to be addressed
*update depending on ability to bind indicators with outcomes*

@@

In addition to the inequality and indicator predictors provided with the dataset, I would like to examine differences in TB outcomes based on level of TB burden. This can be completed by subsetting the high burden countries into a smaller data set.

@@

The data cleaning will consist of widening the data set so each TB indicator is placed in a column in order to explore the indicators separately. The data set will also be split into a full data set containing all countries (wide_data) and high burden countries (highburden). The data will then be summarized and described using exploratory analysis approaches such as graphs, plots, and tables. The statistical analysis will use cross-validation and decision tree modeling to model TB outcomes based on the available predictor variables. A similar approach will be used for the health equity indicators as outcomes.

@@

For the statistical analysis of the data, I created decision tree models for the main outcomes (TB incidence, TB mortality, and TB prevalence) based on the subgroups of the indicator and tuned the models using cross-validation folds. The data was initially split into test/train data sets to better evaluate model performance. An example is shown below. This method was repeated for the `wide_data` and `highburden` countries.

@@

Other indicators were also analyzed and fit to a decision tree model using the same methods.

@@

I would still like to try to manipulate the data set so I can use the social determinant indicators as predictors for the main outcomes. This will most likely be limited to the high burden countries.

@@

Please see `exploratory_analysis.qmd` for full code describing data exploratory analysis.

@@

Most indicators showed the greatest disparities due to economic status and education level.

## Basic statistical analysis
*update depending on binding outcomes and predictors: will probably do linear model predictions from TB outcomes from one indicator if possible. otherwise will do line*

@@

For the majority of the statistical analysis, the tuberculosis outcomes and the health equity indicators with high disparity between subgroups (as identified during the exploratory analysis) were modeled and tuned using the decision tree model and cross-validation methods from the `tidymodels` guide. The models were all used the training data to tune the cost_complexity and tree_depth parameters to find the model with the lowest value of RMSE. An example output is shown below.

```{r}
#tree plot models for full data TB incidence
readRDS(file = here("results", "tune plots", "wdinctreeplot.rds")) 
#tree plot models for high burden TB incidence
readRDS(file = here("results", "tune plots", "hbinctreeplot.rds"))
```

Please refer to the `statistical_analysis.qmd` for the full code and results.

A summary of the model performances are shown below. The full data set models are marked with 'wd' and the high burden models are marked as 'hb'. The outcomes being modeled are abbreviated in the `model` column. The best performing model predicted prevalence to notification ratio for the full data set (RMSE = 0.971). The TB outcome models all performed poorly in predicting TB incidence, mortality, and prevalence. This is most likely due to the only significant predictor being country.
```{r}
readRDS(file = here("results", "summaryrmse.rds")) #summary table of RMSE from decision tree models
```

The second research question was to explore the differences between the full data set and high burden countries. I hypothesized that the high burden models would have a higher performance because these countries have the highest concentration of TB cases. However, the high burden models consistently preformed worse than the full data models. This is most likely due to the lower amount of available data. 
{{< pagebreak >}}
